{fairmetrics}: An R package for group fairness evaluation

### 2025-05-30 paper.Rmd
Edit Statement of Need.

<!--- 
Rewritten by Jesse
--->
Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for certain groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of model development, as biased models can perpetuate or exacerbate existing social inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). These criteria assess whether a model is equally accurate or well-calibrated across predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions, and includes example a dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

<!---
Jianhui - this is too similar to the original paper still and I suggest we emphasize that the package isn't restricted to biomedical data.  We can of course use it as an example but this isn't the need for the package. Can you rewrite this to just be about the general use of fairness?  i.e., point 1 - ml increasing integrated into different areas, point 2 - give a few examples, point 3 - need for the package in light of other work.  I don't think interpetable is the correct word here, the key difference is allowing for uncertainty quantification, we also provide ratio + difference estimates, i dont think R fairness package does, I would emphasize   
--->

Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21]. As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity. For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants[@mattuMachineBias]. As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles. Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients[@obermeyerDissectingRacialBias2019]. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients. Therefore, there is a urgent need for companies, engineers, and researchers to ensure that ML models are fair and equitable across different groups before they are deployed in real-world applications.

While existing software can compute fairness metrics, they provide only point estimates without quantifying the uncertainty around those metrics. This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations. The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, enabling users to make statistically grounded decisions about the fairness of their models.

>>>>

Fairness is a growing area of machine learning (ML) that focuses on ensuring models do not produce systematically biased outcomes for particular groups, particularly those defined by protected attributes such as race, gender, or age. Evaluating fairness is a critical aspect of ML model development, as biased models can perpetuate structural inequalities.  The {fairmetrics} R package offers a user-friendly framework for rigorously evaluating numerous group-based fairness criteria, including metrics based on independence (e.g., statistical parity), separation (e.g., equalized odds), and sufficiency (e.g., predictive parity). Group-based fairness criteria assess whether a model is equally accurate or well-calibrated across a set of predefined groups so that appropriate bias mitigation strategies can be implemented. {fairmetrics} provides both point and interval estimates for multiple metrics through convenient wrapper functions and includes an example dataset derived from the Medical Information Mart for Intensive Care, version II (MIMIC-II) database [@goldberger2000physiobank; @raffa2016clinical].

@@

ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21].  Mounting evidence suggest that these models often exhibit bias across groups defined by protected attributes. For example, within criminal justice, the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) software, a tool used by U.S. courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants[@mattuMachineBias].  This bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to white defendants with similar risk profiles. Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S. to identify patients with complex health needs for high-risk care management programs was shown to be significantly lower calibrated for Black patients relative to white patients[@obermeyerDissectingRacialBias2019]. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.  These examples illustrate that there is an urgent need for ML practitioners and researchers to ensure that ML models support fair decision before they are deployed in real-world applications.

While existing software can compute group fairness criteria, they only point estimates and/or visualizations without quantifying the uncertainty around the criteria. This limitation prevents users from determining whether observed disparities between groups are statistically significant or merely the result of random variation due to finite sample size, potentially leading to incorrect conclusions about fairness violations. The {fairmetrics} R package addresses this gap by providing bootstrap-based confidence intervals (CIs) for both difference-based and ratio-based group fairness metrics, empowering users to make statistically grounded decisions about the fairness of their models, which is inconsistently done in practice.
